Follow-up Samples

### 2017-02-16 submission/manuscript_outline_20161024.tex

Colorectal cancer (CRC) continues to be a leading cause of cancer
related deaths and is currenlty the third most common cause of cancer
deaths {[}1,2{]}. Over the last few years death due to the disease has
seen a significant decrease, thanks mainly to improvements in screening
{[}1{]}. However, despite this improvement there are still approximately
50,000 deaths from the disease a year {[}2{]}.

Recently, there has been promising work on the bacterial microbiome and
it's ability to be able to complement existing screening methods such as
Fecal Immunoglobulin Test (FIT) or act alone as a screening tool
{[}3,4{]}. There has also been research into how this microbiome could
be altered directly on tumor tissue itself {[}5{]}. A few studies have
also shown how this microbiome {[}6{]} or specific members within it
{[}7{]} could be directly involved with the pathogenesis of CRC. These
studies have helped to provide a tantalizing link between the bacterial
microbiome and CRC. Although these studies suggest that the bacterial
microbiome might change after treatment there remains limited
information on the bacterial microbiome before and after surgerical
removal of lesion (adenoma or carcinoma) and whether the community
changes at all.

In this study we tested the hypothesis that the bacterial microbiome
changes after surgery for individuals with a lesion. Our analysis
included both alpha and beta diversity analysis along with investigation
of individual operational taxonomic units (OTUs). We also utilized
Random Forest models and observed how these models as well as specific
OTUs within them performed pre (initial) and post (follow up) surgery.
We also used these models to look for similar important OTUs to identify
the crucial OTUs for not only classifying initial and follow up samples
but also lesion or normal.

@@

\textbf{\emph{Positive Probability after Lesion Removal:}} If there were
common OTUs for individuals with adenoma and carcinoma, that were
different versus normal controls, we would expect to find a decrease in
the positive probability of the follow up sample to be either lesion or
an initial sample. This is what we observed regardless of model used
(lesion or initial sample) or whether it was built on the full or
reduced variable data set {[}Figure 4 \& S4{]} (P-value \textless{}
0.001).

When we seperated individuals based on whether they had an adenoma or
carcinoma there was only a decrease in positive probability for the
carcinoma group (P-value \textless{} 0.001) and not for the adenoma
group (P-value \textgreater{} 0.05). We also observed that there were no
significant differences in whether the models classified the samples as
having lesion between the predicted and actual (P-value \textgreater{}
0.05). This lack of difference between the predicted and actual
classifications were also observed for the initial sample model (P-value
\textgreater{} 0.05). Even though the lesion model was not as accurate
in classifying samples as the initial sample model. It was able to
correctly keep the one individual who still had a carcinoma on follow up
above the cut off threshold for a positive call {[}Figure 4a \& S4a{]}
while the initial sample model did not {[}Figure 4b \& S4b{]}.

@@

From our results there were some large observed differences in the
bacterial microbiome between pre and post surgery samples based on
whether the individual had an adenoma or carcinoma. There was much
larger differences between initial and follow up samples based on the
thetayc distance metric and in fecal blood as measured by FIT for
individuals with carcinoma versus adenoma {[}Figure 1{]}. However, there
were no differences between initial and follow up samples for Shannon
Diversity, observed OTUs, or evenness regardless of whether the
individual had an adenoma or carcinoma {[}Table S1{]}. There was also no
differences in relative abundance of any specific OTU for lesion,
adenoma only, or carcinoma only {[}Figure s1{]}.

@@

We next created a model that incorporated FIT and the bacterial
microbiome to either be able to classify lesions (adenoma or carcinoma)
or initial samples in order to find common OTUs in the community that
change for both adenoma and carcinoma. What we found was that the
commonly associated CRC bacteria were not higly represented within our
models but rather that OTUs that made up the most important variables
overwhelmingly belonged to commensal bacteria. With only the lesion
model having a single OTU from a previously associated cancer bacterium
(\emph{Porphyromonas asaccharolytica}). Using only these important OTUs
and FIT both models (lesion and before sample) significantly decreased
positive probability of either lesion or being an initial sample on
follow up {[}Figure 4 \& S4{]}. Further confirmation of the importance
of the changes of commensal bacteria to these classifications was that a
total of 23 OTUs were common to both models and the vast majority
belonged to regular residents of our gut community.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are a number of reasons why this may
have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that it is possible that our Random
Forest models were able to gather the same information from measures
such as FIT or other OTUs. It is also possible that both of these
explanations could have played a role. Regardless, our observations
would suggest that an individual's resident bacteria have a large role
to play in disease initiation and could change in a way that allows
predictive models to lower the positive probability of a lesion after
surgery {[}Figure 4{]}. It should be noted that our study does not argue
against the importance of these CRC associated bacteria in the
pathogenesis of disease but rather that the models do not utilize these
specific bacteria for classification purposes (lesion or before sample).
In fact, it is possible that these CRC associated bacteria are important
in the transition from adenoma to carcinoma and would be one explanation
as to why in our data we not only see high initial relative abundances,
in certain individuals, but also large decreases in relative abundance
in those with carcinoma but not in those with adenoma after surgery
{[}Figure 2{]}.

Many of the common OTUs between the different models used had many OTUs
that taxonomically classifed to potential butyrate producers {[}Table
S2{]}. Another batch of OTUs classifed to bacteria that can either
degrade polyphenols or are inhibited by them. Both butyrate and
polyphenols are thought to protective against cancer in part by reducing
inflammation {[}10{]}. These protective compounds are derived from the
breakdown of fiber, fruits, and vegetables by resident gut microbes. One
example of this potential diet-microbiome-inflammation-polyp axis is
that \emph{Bacteroides}, which was highly prevalent in our models, are
known to be increased in those with high non-meat based protein
consumption {[}11{]}. High protein consumption in general has been
linked with an increased CRC risk {[}12{]}. Conversely,
\emph{Bacteroides} are inhibited by polyphenols which are derived from
fruits and vegetables {[}13{]}. Our data fits with the hypothesis that
the microbial metabolites from breakdown products within our own diet
could not only help to shape the exisiting community but also have an
effect on CRC risk and disease progression.

One limitation of our study is that we do not know whether individuals
who were still classified as positive by the lesion model eventually had
a subsequent CRC diagnosis. This information would help to strengthen
the case for our lesion model to have kept a number of individuals above
the cutoff threshold even though at follow up they were diagnosed as no
longer having a lesion. Another limitation is that we do not know if
adding modern tests such as the stool DNA test {[}14{]} could help
improve our overall AUC. Another limitation is that this study drew
heavily from those with Caucasian ancestry. The results may not be
immediately representative of those with either Asian or African
ancestry. Finally, although our training and test set are relatively
large we still run the risk of over-fitting or having a model that may
not be immediately extrapolate-able to other populations. We've done our
best to safeguard against this by not only running 10-fold cross
validation but also having over 100 different 80/20 splits to try and
mimic the type of variation that might be expected to occur. The time
difference in collection of sample between adenoma and carcinoma could
have affected our observed results for the differences between
individuals with adenoma or carcinoma. This confounding though does not
affect the observations based on the overall lesion results.

Another interesting outcome was that within figure 3 the before sample
model showed better test AUC results then the training set AUC. This may
have occurred because the training AUC that was determined from 20
repeated 10 fold cross validation removed samples at random and did not
take into account that they were matched samples. Another potential
reason is that the model itself may be over-fit since the total number
of samples was not that large. However, the lesion model did not suffer
from these discrepancies and similar conclusions can be drawn solely
from this model. Regardless, further independent studies will need to be
carried out to verify our findings since not only are we dealing with
feces, which could be very different than the communities present on the
actual tissue, but also are dealing with correlations that may not be
representative of the true pathogenesis of disease.

Despite these limitations we think that these findings significantly add
to the existing scientific knowledge on CRC and the bacterial
microbiome: That there is a measurable difference in the bacterial
community after surgical removal of lesion. Further, the ability for
machine learning algorithms to take bacterial microbiome data and
successfully lower positive probability after either adenoma or
carcinoma removal provides evidence that there are specific signatures,
mostly attributable to commensal organisms, associated with these
lesions. Our data provides evidence that commensal bacteria are
important in the development of polyps and also potentially the
transition from adenoma to carcinoma.

@@

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al
{[}3{]}. In brief, study exclusion involved those who had already
undergone surgery, radiation, or chemotherapy, had colorectal cancer
before a baseline fecal sample could be obtained, had IBD, a known
hereditary non-polyposis colorectal cancer, or Familial adenomatous
polyposis. Samples used to build the models for prediction were
collected either prior to a colonoscopy or between 1 - 2 weeks after.
The bacterial microbiome has been shown to normalize within this time
period {[}15{]}. Our follow up data set had a total of 67 individuals
that not only had a sample as described but also a follow up sample
between 188 - 546 days after surgery and treatment had been completed.
This study was approved by the University of Michigan Institutional
Review Board. All study participants provided informed consent and the
study itself conformed to the guidelines set out by the Helsinki
Declaration.

\textbf{\emph{FIT and 16S rRNA Gene Sequencing}} FIT was analyzed as
previously published using both OC FIT-CHEK and OC-Auto Micro 80
automated system (Polymedco Inc.) {[}16{]}. 16S rRNA gene sequencing was
completed as previously described by Kozich, et al. {[}17{]}. In brief,
DNA extraction used the 96 well Soil DNA isolation kit (MO BIO
Laboratories) and an epMotion 5075 automated pipetting system
(Eppendorf). The V4 variable region was amplified and the resulting
product was split between three sequencing runs with normal, adenoma,
and carcinoma evenly represented on each run. Each group was randomly
assigned to avoid biases based on sample collection location.

\textbf{\emph{Sequence Processing}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described {[}17{]}. The general processing workflow
using mothur was as follows: Paired-end reads were first merged into
contigs, quality filtered, aligned to the SILVA database, screened for
chimeras, classified with a naive Bayesian classifier using the
Ribosomal Database Project (RDP), and clustered into Operational
Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
neighbor clustering algorithm. The number of sequences for each sample
was rarefied to 10523 in an attempt to minimize uneven sampling.

\textbf{\emph{Lesion Model Creation}} The Random Forest {[}18{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main testing and training of the model
completed on a data set of 490 individuals. This model was then applied
to our follow up data set of 67 individuals. The model included data on
FIT and the bacterial microbiome. Non-binary data was checked for near
zero variance and OTUs that had near zero variance were removed. This
pre-processing was performed with the R package caret (v6.0.73).
Optimization of the mtry hyper-parameter involved taking the samples and
making 100 different 80/20 (train/test) splits of the data where normal
and lesion were represented in the same proportion within both the whole
data set and the 80/20 split. Each of these splits were then run through
20 repeated 10-fold cross validations to optimize the mtry
hyper-parameter by maximizing the AUC (Area Under the Curve of the
Receiver Operator Characteristic). This resulting model was then tested
on the 20\% of the data that was originally held out from this overall
process. Once the ideal mtry was found the entire 490 sample set was
used to create the final Random Forest model on which classifications on
the 67-person cohort was completed. The default cutoff of 0.5 was used
as the threshold to classify individuals as positive or negative for
lesion. The hyper-parameter, mtry, defines the number of variables to
investigate at each split before a new division of the data is created
with the Random Forest model.

\textbf{\emph{Before Sample Model Creation}} We also investigated
whether a model could be created that could identify before and after
surgery samples. The main difference was that only the 67-person cohort
was used at all stages of model building and classification. Other than
this difference the creation of this model and optimization of the mtry
hyper-parameter was completed using the same procedure that was used to
create the lesion model. Instead of classifying samples as positive or
negative of lesion this model classified samples as positive or negative
for being a before surgery sample.

\textbf{\emph{Selection of Important OTUs}} In order to assess which
variables were most important to all the models we counted the number of
times a variable was present in the top 10\% of mean decrease in
accuracy (MDA) for each of the 100 different 80/20 split models and then
filtered this list to variables that were only present more than 50\% of
the time. This final collated list of variables was what was considered
the most important for the lesion or before sample models.

\textbf{\emph{Statistical Analysis}} The R software package (v3.3.2) was
used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}19{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between initial and follow up samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was needed
a Benjamini-Hochberg (BH) correction was applied {[}20{]} and a
corrected P-value of less than 0.05 was considered significant. Unless
otherwise stated the P-values reported are those that were BH corrected.

\textbf{\emph{Analysis Overview}} Initial and follow up samples were
analyzed for differences in alpha and beta diversity. Next, differences
in FIT between initial and follow ups for either adenoma or carcinoma
were investigated. From here, all OTUs that were used in either model
were then analyzed using a paired Wilcoxson test. We then investigated
the relative abundance of specific previously associated CRC bacteria,
specifically, OTUs that taxonomically classified to \emph{Fusobacterium
nucleatum}, \emph{Parvimonas micra}, \emph{Peptostreptococcus
assacharolytica}, and \emph{Porphyromonas stomatis}. We wanted to test
if there were any differences based on whether the individual had an
adenoma or carcinoma. From here the lesion model was then tested for
accuracy in prediction and whether it reduced the positive probability
of lesion after surgery. The most important OTUs for this were used to
build a reduced model and it was assessed for similarity to the original
model. We then used the before sample model to assess whether it could
classify samples better then the lesion model. The most important OTUs
were then identified from this model and used to create a reduced
feature before sample model. This reduced feature model, as was done
with the lesion model, was compared to the full model for loss of
accuracy. Finally, a list of common OTUs were found for the two
different models used.

\textbf{\emph{Reproducible Methods.}} A detailed and reproducible
description of how the data were processed and analyzed can be found at
\url{https://github.com/SchlossLab/Sze_followUps_2017}. Raw sequences
have been deposited into the NCBI Sequence Read Archive (SRP062005 and
SRP096978) and the necessary metadata can be found at
\url{https://www.ncbi.nlm.nih.gov/Traces/study/} and searching the
respective SRA study accession.

@@

\textbf{Figure 1: General Differences between the Adenoma or Carcinoma
Group.} A) A significant difference was found between the adenoma and
carcinoma group for thetayc (P-value = 0.000472). B) A significant
difference was found between the adenoma and carcinoma group for change
in FIT (P-value = 2.15e-05). C) NMDS of the initial and follow up
samples for the Adenoma group. D) NMDS of the initial and follow up
samples for the Carcinoma group. For C) and D) the teal represents
initial samples and the pink represents follow up samples.

\textbf{Figure 2: Previously Associated CRC Bacteria in Initial and
Follow up Samples.} A) Carcinoma initial and follow up samples. There
was a significant difference in initial and follow up sample for the
OTUs classified as \emph{Peptostreptococcus stomatis} (P-value = 0.0496)
and \emph{Porphyromonas asaccharolytica} (P-value = 0.00842). B) Adenoma
initial and follow up samples. There were no significant differences
between initial and follow up.

\textbf{Figure 3: Graph of the Receiver Operating Characteristic Curve
for lesion and Before Sample Models.} The shaded areas represents the
range of values of a 100 different 80/20 splits of the test set data
using either all variables (grey) or reduced variable (red) models. The
blue line represents the reduced variable model using 100\% of the data
set. A) Lesion model. B) Before sample model

\textbf{Figure 4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Before Sample Reduced Variable Models} A) Lesion
positive probability adjustment of those with carcinoma from initial to
follow up sample B) Initial follow up positive probability adjustment of
those with carcinoma from initial to follow up sample C) Lesion positive
probability adjustment of those with adenoma as well as those with SRN
and the probability adjustment from initial to follow up sample. D)
Initial follow up positive probability adjustment of those with adenoma
as well as those with SRN and the probability adjustment from initial to
follow up sample. The dotted line represents the threshold used to make
the decision of whether a sample was positive or not.

@@

\textbf{Figure S1: Distribution of P-values from Paired Wilcoxson
Analysis of OTUs in Initial versus Follow Up}

@@

\textbf{Figure S4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Before Sample Full Variable Models} A) Lesion
positive probability adjustment of those with carcinoma from initial to
follow up sample B) Initial follow up positive probability adjustment of
those with carcinoma from initial to follow up sample C) Lesion positive
probability adjustment of those with adenoma as well as those with SRN
and the probability adjustment from initial to follow up sample. D)
Initial follow up positive probability adjustment of those with adenoma
as well as those with SRN and the probability adjustment from initial to
follow up sample. The dotted line represents the threshold used to make
the decision of whether a sample was positive or not.

\textbf{Figure S5: Thetayc Graphed Against Time of Follow up Sample from
Initial}

>>>>

Colorectal cancer (CRC) continues to be a leading cause of cancer
related deaths and is currently the third most common cause of cancer
deaths {[}1,2{]}. Over the last few years death due to the disease has
seen a significant decrease, thanks mainly to improvements in screening
{[}1{]}. However, despite this improvement there are still approximately
50,000 deaths from the disease a year {[}2{]}.

Recently, there has been promising work on the bacterial microbiome and
it's ability to be able to complement existing screening methods such as
Fecal Immunoglobulin Test (FIT) or act alone as a screening tool
{[}3,4{]}. There has also been research into how this microbiome could
be altered directly on tumor tissue itself {[}5{]}. A few studies have
also shown how this microbiome {[}6{]} or specific members within it
{[}7{]} could be directly involved with the pathogenesis of CRC. These
studies have helped to provide a tantalizing link between the bacterial
microbiome and CRC. Although these studies suggest that the bacterial
microbiome might change after treatment there remains limited
information on the bacterial microbiome before and after surgical
removal of lesion (adenoma or carcinoma) and whether the community
changes at all.

In this study we tested the hypothesis that the bacterial microbiome
changes between pre (initial) and post (follow up) samples after surgery
for individuals with a lesion (adenoma or carcinoma). Our analysis
included both alpha and beta diversity analysis along with investigation
of individual operational taxonomic units (OTUs). We also utilized
Random Forest models and observed how these models, as well as specific
OTUs within them, performed pre and post surgery. We also investigated
the models for similar important OTUs to identify the crucial OTUs for
not only classifying initial and follow up samples but also lesion or
normal.

@@

\textbf{\emph{Positive Probability after Lesion Removal:}} If there were
common OTUs that could separate adenoma and carcinoma from normal
controls, we would expect to find a decrease in the positive probability
of the follow up sample to be either lesion or an initial sample. This
is what we observed regardless of model used (lesion or initial sample)
or whether it was built on the full or reduced variable data set
{[}Figure 4 \& S4{]} (P-value \textless{} 0.001).

When we separated individuals based on whether they had an adenoma or
carcinoma there was only a decrease in positive probability for the
carcinoma group (P-value \textless{} 0.001) and not for the adenoma
group (P-value \textgreater{} 0.05). We also observed that there were no
significant differences in whether the models classified the samples as
having lesion between the predicted and actual (P-value \textgreater{}
0.05). This lack of difference between the predicted and actual
classifications were also observed for the initial sample model (P-value
\textgreater{} 0.05). Even though the lesion model was not as accurate
in classifying samples as the initial sample model. It was able to
correctly classify the one individual who still had a carcinoma on
follow up {[}Figure 4a \& S4a{]} while the initial sample model did not
{[}Figure 4b \& S4b{]}.

@@

From our results there were large observed differences in the bacterial
microbiome between pre and post surgery samples based on whether the
individual had an adenoma or carcinoma. There were much larger
differences between initial and follow up samples based on the thetayc
distance metric and in fecal blood as measured by FIT for individuals
with carcinoma versus adenoma {[}Figure 1{]}. However, there were no
differences between initial and follow up samples for Shannon Diversity,
observed OTUs, or evenness regardless of whether the individual had an
adenoma or carcinoma {[}Table S1{]}. There was also no differences in
relative abundance of any specific OTU for lesion, adenoma only, or
carcinoma only {[}Figure s1{]}. Although there was a detectable change
in the carcinoma group. It should be noted that whether this change was
towards what wouldbe be expected for normal controls is not known.

@@

We next created a model that incorporated FIT and the bacterial
microbiome to either be able to classify lesions (adenoma or carcinoma)
or initial samples in order to find common OTUs in the community that
change for both adenoma and carcinoma. What we found was that the
commonly associated CRC bacteria were not highly represented within our
models but rather that OTUs that made up the most important variables
overwhelmingly belonged to commensal bacteria. With only the lesion
model having a single OTU from a previously associated cancer bacterium
(\emph{Porphyromonas asaccharolytica}). Using only these important OTUs
and FIT, both models (lesion and initial sample) significantly decreased
positive probability of either lesion or being an initial sample on
follow up {[}Figure 4 \& S4{]}. Further confirmation of the importance
of the changes of commensal bacteria to these classifications was that a
total of 23 OTUs were common to both models and the vast majority
belonged to regular residents of our gut community {[}Table S3{]}.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are a number of reasons why this may
have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from measures such as FIT or other
OTUs. It is also possible that both of these explanations could have
played a role. Regardless, our observations would suggest that an
individual's resident bacteria have a large role to play in disease
initiation and could change in a way that allows predictive models to
lower the positive probability of a lesion after surgery {[}Figure 4{]}.
It should be noted that our study does not argue against the importance
of these CRC associated bacteria in the pathogenesis of disease but
rather that they are not the main bacteria changing after surgical
removal of lesion. In fact, it is possible that these CRC associated
bacteria are important in the transition from adenoma to carcinoma and
would be one explanation as to why in our data we not only see high
initial relative abundances in carcinoma and not adenoma individuals but
also large decreases in relative abundance in some of those with
carcinoma but not in those with adenoma after surgery {[}Figure 2{]}.

Many of the common OTUs between the different models used had many OTUs
that taxonomically classified to potential butyrate producers {[}Table
S2{]}. Another batch of OTUs classified to bacteria that can either
degrade polyphenols or are inhibited by them. Both butyrate and
polyphenols are thought to be protective against cancer in part by
reducing inflammation {[}10{]}. These protective compounds are derived
from the breakdown of fiber, fruits, and vegetables by resident gut
microbes. One example of this potential
diet-microbiome-inflammation-polyp axis is that \emph{Bacteroides},
which was highly prevalent in our models, are known to be increased in
those with high non-meat based protein consumption {[}11{]}. High
protein consumption in general has been linked with an increased CRC
risk {[}12{]}. Conversely, \emph{Bacteroides} are inhibited by
polyphenols which are derived from fruits and vegetables {[}13{]}. Our
data fits with the hypothesis that the microbial metabolites from
breakdown products within our own diet could not only help to shape the
existing community but also have an effect on CRC risk and disease
progression.

One limitation of our study is that we do not know whether individuals
who were still classified as positive by the lesion model eventually had
a subsequent CRC diagnosis. This information would help to strengthen
the case for our lesion model keeping a number of individuals above the
cutoff threshold even though at follow up they were diagnosed as no
longer having a lesion. Another limitation is that we do not know if
adding modern tests such as the stool DNA test {[}14{]} could help
improve our overall AUC. This study also drew heavily from those with
Caucasian ancestry making it possible that the observations may not be
representative of those with either Asian or African ancestry. Although
our training and test set are relatively large we still run the risk of
over-fitting or having a model that may not be representative of other
populations. We've done our best to safeguard against this by not only
running 10-fold cross validation but also having over 100 different
80/20 splits to try and mimic the type of variation that might be
expected to occur. The time difference in collection of sample between
adenoma and carcinoma could have affected our observations for
differences between individuals with adenoma or carcinoma. This
confounding though would not affect the observations where these
individuals are grouped together.

Interestingly, within figure 3 the before sample model showed better
test AUC results then the training set AUC. This may have occurred
because the training AUC determined from 20 repeated 10 fold cross
validation removed samples at random and did not take into account that
they were matched samples. Another potential reason is that the model
itself may be over-fit since the total number of samples was not that
large. However, the lesion model did not suffer from these discrepancies
and similar conclusions can be drawn solely from this model alone.
Further independent studies need to be carried out to verify our
findings since not only are we dealing with feces, which could be very
different than the communities present on the actual tissue, but also
are dealing with correlations that may not be representative of the true
pathogenesis of disease.

Despite these limitations our findings add to the existing scientific
knowledge on CRC and the bacterial microbiome: That there is a
measurable difference in the bacterial community after surgical removal
of lesion. Further, the ability for machine learning algorithms to take
bacterial microbiome data and successfully lower positive probability
after either adenoma or carcinoma removal provides evidence that there
are specific signatures, mostly attributable to commensal organisms,
associated with these lesions. Our data provides evidence that commensal
bacteria may be important in the development of polyps and also
potentially the transition from adenoma to carcinoma.

@@

\textbf{\emph{Study Design and Patient Sampling:}} The sampling and
design were similar to that reported in Baxter, et al {[}3{]}. In brief,
study exclusion involved those who had already undergone surgery,
radiation, or chemotherapy, had colorectal cancer before a baseline
fecal sample could be obtained, had IBD, a known hereditary
non-polyposis colorectal cancer, or Familial adenomatous polyposis.
Samples used to build the models for prediction were collected either
prior to a colonoscopy or between 1 - 2 weeks after. The bacterial
microbiome has been shown to normalize within this time period {[}15{]}.
Our follow up data set had a total of 67 individuals that not only had a
sample as described but also a follow up sample between 188 - 546 days
after surgery and treatment had been completed. This study was approved
by the University of Michigan Institutional Review Board. All study
participants provided informed consent and the study itself conformed to
the guidelines set out by the Helsinki Declaration.

\textbf{\emph{FIT and 16S rRNA Gene Sequencing:}} FIT was analyzed as
previously published using both OC FIT-CHEK and OC-Auto Micro 80
automated system (Polymedco Inc.) {[}16{]}. 16S rRNA gene sequencing was
completed as previously described by Kozich, et al. {[}17{]}. DNA
extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories)
and an epMotion 5075 automated pipetting system (Eppendorf). The V4
variable region was amplified and the resulting product was split
between three sequencing runs with normal, adenoma, and carcinoma evenly
represented on each run. Each group was randomly assigned to avoid
biases based on sample collection location.

\textbf{\emph{Sequence Processing:}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described {[}17{]}. The general processing workflow
using mothur was as follows: Paired-end reads were first merged into
contigs, quality filtered, aligned to the SILVA database, screened for
chimeras, classified with a naive Bayesian classifier using the
Ribosomal Database Project (RDP), and clustered into Operational
Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
neighbor clustering algorithm. The number of sequences for each sample
was rarefied to 10523 in an attempt to minimize uneven sampling.

\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}18{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main testing and training of the model
completed on a data set of 423 individuals. This model was then applied
to our follow up data set of 67 individuals. In brief, the model
included data on FIT and the bacterial microbiome. Non-binary data was
checked for near zero variance and OTUs that had near zero variance were
removed. This pre-processing was performed with the R package caret
(v6.0.73). Optimization of the mtry hyper-parameter involved taking the
samples and making 100 different 80/20 (train/test) splits of the data
where normal and lesion were represented in the same proportion within
both the whole data set and the 80/20 split. Each of these splits were
then run through 20 repeated 10-fold cross validations to optimize the
mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the
Receiver Operator Characteristic). This resulting model was then tested
on the 20\% of the data that was originally held out from this overall
process. Once the ideal mtry was found the entire 423 sample set was
used to create the final Random Forest model on which classifications on
the 67-person cohort was completed. The default cutoff of 0.5 was used
as the threshold to classify individuals as positive or negative for
lesion. The hyper-parameter, mtry, defines the number of variables to
investigate at each split before a new division of the data is created
with the Random Forest model.

\textbf{\emph{Initial Sample Model Creation:}} We also investigated
whether a model could be created that could identify pre (initial) and
post (follow up) surgery samples. The main difference was that only the
67-person cohort was used at all stages of model building and
classification. Other than this difference the creation of this model
and optimization of the mtry hyper-parameter was completed using the
same procedure as was used for the lesion model. Instead of classifying
samples as positive or negative of lesion this model classified samples
as positive or negative for being an initial surgery sample.

\textbf{\emph{Selection of Important OTUs:}} In order to assess which
variables were most important to either model we counted the number of
times a variable was present in the top 10\% of mean decrease in
accuracy (MDA) for each of the 100 different 80/20 split models and then
filtered this list to variables that were only present more than 50\% of
the time. This final collated list of variables was what was considered
the most important for the lesion or before sample models.

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}19{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between initial and follow up samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate a Benjamini-Hochberg (BH) correction was applied {[}20{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected.

\textbf{\emph{Analysis Overview:}} We first wanted to test if there were
any differences based on whether the individual had an adenoma or
carcinoma. This was done by testing initial and follow up samples for
differences in alpha and beta diversity, testing differences in FIT
between initial and follow ups, testing all OTUs that were used in
either built model, and investigating the relative abundance of specific
previously associated CRC bacteria (\emph{Fusobacterium nucleatum},
\emph{Parvimonas micra}, \emph{Peptostreptococcus assacharolytica}, and
\emph{Porphyromonas stomatis}) based on adenoma and carcinoma. From here
the lesion model was then tested for accuracy in prediction and whether
it reduced the positive probability of lesion after surgery. The most
important OTUs for this were used to build a reduced model and it was
assessed for similarity to the original model. We then used the initial
sample model to assess whether it could classify samples better then the
lesion model. The most important OTUs were then identified from this
model and used to create a reduced feature initial sample model. This
reduced feature model, as was done with the lesion model, was compared
to the full model for loss of accuracy. A list of common OTUs were found
for the two different models used. Finally, both lesion and initial
sample models initial positive probabilites were tested against the
follow up positive probabilities.

\textbf{\emph{Reproducible Methods:}} A detailed and reproducible
description of how the data were processed and analyzed can be found at
\url{https://github.com/SchlossLab/Sze_followUps_2017}. Raw sequences
have been deposited into the NCBI Sequence Read Archive (SRP062005 and
SRP096978) and the necessary metadata can be found at
\url{https://www.ncbi.nlm.nih.gov/Traces/study/} and searching the
respective SRA study accession.

@@

\textbf{Figure 1: General Differences between the Adenoma or Carcinoma
Group.} A) A significant difference was found between the adenoma and
carcinoma group for thetayc (P-value = 0.000472). B) A significant
difference was found between the adenoma and carcinoma group for change
in FIT measurement (P-value = 2.15e-05). C) NMDS of the initial and
follow up samples for the Adenoma group. D) NMDS of the initial and
follow up samples for the Carcinoma group. For C) and D) the teal
represents initial samples and the pink represents follow up samples.

\textbf{Figure 2: Previously Associated CRC Bacteria in Initial and
Follow up Samples.} A) Carcinoma initial and follow up samples had an
observed significant difference in initial and follow up sample for the
OTUs classified as \emph{Parvimonas micra} (P-value = 0.0116) and
\emph{Porphyromonas asaccharolytica} (P-value = 0.00842). B) Adenoma
initial and follow up samples. There were no significant differences
between initial and follow up.

\textbf{Figure 3: Receiver Operating Characteristic Curve for Lesion and
Initial Sample Models.} The shaded areas represents the range of values
of a 100 different 80/20 splits of the test set data using either all
variables (grey) or reduced variable (red) models. The blue line
represents the reduced variable model using 100\% of the data set and
what was used for subsequent classification. A) Lesion model. B) Before
sample model

\textbf{Figure 4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Initial Sample Reduced Variable Models} A) Lesion
positive probability change from initial to follow up sample in those
with carcinoma B) Initial positive probability change from initial to
follow up sample in those with carcinoma C) Lesion positive probability
change from initial to follow up sample if those with adenoma or Screen
Relevent Neoplasia (SRN). D) Initial positive probability change from
initial to follow up sample in those with adenoma or SRN. The dotted
line represents the threshold used to make the decision of whether a
sample was positive or not.

@@

\textbf{Figure S1: Distribution of P-values from Paired Wilcoxson
Analysis of All OTUs for Initial versus Follow Up}

@@

\textbf{Figure S4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Before Sample Full Variable Models} A) Lesion
positive probability change from initial to follow up sample in those
with carcinoma B) Initial positive probability change from initial to
follow up sample in those with carcinoma C) Lesion positive probability
change from initial to follow up sample in those with adenoma or Screen
Relevent Neoplasia (SRN). D) Initial positive probability change from
initial to follow up sample in those with adenoma or SRN. The dotted
line represents the threshold used to make the decision of whether a
sample was positive or not.

\textbf{Figure S5: Thetayc Versus Time of Follow up Sample from Initial}
